On the reality of the quantum state once again: A no-go theorem for $\psi$-ontic models

### 2023-02-13 Articles/2021-HarriganSpekkens/HarriganSpekkensTakeTwo.tex
HarriganSpekkens: adding failure composite systems

Combining our result with the PBR argument, we conclude that the HS classification itself is fundamentally flawed: both $\psi$-epistemic and $\psi$-ontic models contradict quantum mechanics.\footnote{In principle, PBR leaves open the possibility of $\psi$-epistemic models that violate statistical independence. Given that statistical independence is linked to entropy additivity, that loophole can be closed as well.} Consequently, the HS framework should be employed neither to classify interpretations of the quantum formalism, nor to draw conclusions about the nature of $\psi$.

@@

If we combine this result with PBR, the only models that are not ruled out are $\psi$-epistemic models where the objects are not independently prepared. That is, where the joint distribution over ontic states is not the product of the marginals. However, this loophole can also be closed with information theoretic considerations. It is a well-known result in both classical and quantum information theory that the entropy of a joint distribution is the sum of the marginal if and only if the subsystems are independent (\cite{Ash:2010}, \cite{Nielsen:2010}). Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This tells us that $\psi$-epistemic models must represent composite states of independently prepared systems with independently distributed epistemic states, and therefore PBR applies. This closes the last loophole: no model in the HS categorization can reproduce quantum theory in its entirety, which includes quantum information and quantum statistical mechanics.

Thus, one concludes the following:
\begin{quote}
	\textbf{No ontological model can reproduce quantum theory}.
\end{quote}
Note that HS do not claim that all interpretations must adhere to their framework, stating (\cite{Harrigan:2010}, p. 134), ``Therefore, to categorize any given interpretation of the quantum formalism as $\psi$-ontic or $\psi$-epistemic, it is first necessary to cast it into the mold of an ontological model. If an interpretation resists being so cast, then it cannot be fit into our categorization.'' The result is that a successful interpretation of quantum mechanics must resists being so cast.

@@

Note that our conclusion relies on the use of standard information theoretic tools. In principle, we do not exclude that someone may find a different construction to reproduce the correct entropies. That is, one may give a different definition of entropy to use on the ontic space. We believe this to be highly unlikely as there are, at least, two objective difficulties that cannot be circumvented.

@@

One, naturally, will object that we haven't strictly proven that any of this is impossible or that there is no reasonable way to do it. Granted. However, it is not up to us to show that HS models cannot be made to work outside of the scope of standard information theory. And in any case, we believe we put forth enough evidence that the HS framework has serious flaws which, at the very least, must be understood by those that use it to draw any conclusion about the interpretation of quantum theory and the nature of $\psi$.

@@

Our conclusion is that the PBR theorem finds that $\psi$-epistemic models are not sustainable not because there is a problem with $\psi$-epistemic models in particular, but rather there is an underlying problem in the HS definitions. If understood in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories, an argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.

>>>>

Combining our result with the PBR argument, we conclude that the HS classification itself is fundamentally flawed: both $\psi$-epistemic and $\psi$-ontic models contradict quantum mechanics.\footnote{In principle, PBR leaves open the possibility of $\psi$-epistemic models that violate statistical independence. Given that statistical independence is linked to entropy additivity, we will see that that loophole can be closed as well.} Consequently, the HS framework should be employed neither to classify interpretations of the quantum formalism, nor to draw conclusions about the nature of $\psi$.

@@

@@

Note that our conclusion relies on the use of standard information theoretic tools. In principle, this does not exclude that someone may find a different construction to reproduce the correct entropies. That is, one may give a different definition of entropy to use on the ontic space. We believe this to be highly unlikely as there are, at least, two objective difficulties that cannot be circumvented.

@@

In other words, the use of standard probability measures to describe the mixing is problematic in the same way that the use of standard probability measures would be a problem to describe the probability of transition. If we step back, we can understand our result and the PBR result in a broader context.

\section{The problem with ontological models}

If we combine our result with PBR, the only models that are not ruled out are $\psi$-epistemic models where the objects are not independently prepared. That is, where the joint distribution over ontic states is not the product of the marginals. However, this loophole can also be closed with information theoretic considerations. It is a well-known result in both classical and quantum information theory that the entropy of a joint distribution is the sum of the marginal if and only if the subsystems are independent (\cite{Ash:2010}, \cite{Nielsen:2010}). Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This tells us that $\psi$-epistemic models must represent composite states of independently prepared systems with independently distributed epistemic states, and therefore PBR applies. This closes the last loophole: no model in the HS categorization can reproduce quantum theory in its entirety, which includes quantum information and quantum statistical mechanics.

Thus, one concludes the following:
\begin{quote}
	\textbf{No ontological model can reproduce quantum theory}.
\end{quote}
Now, the result may seem, at first, preposterous. But once we analyze the mathematics in detail, we see that both results, ours and PBR, are fundamentally finding something, in retrospect, obvious.

It is well established that classical probability theory cannot recover the probability of quantum transitions. So, how does the HS model try to accommodate this? Given that $p(\lambda|P)$ combines using classical probability theory, all the non-classicality must be in $p(k|\lambda, M)$. In fact, note that this second expression does not allow us to write the joint probability distribution for multiple observers; therefore, each ontic state is not partitioned further. Therefore if $\lambda$ is the state of a single quantum system, we can hide the inner product inside $p(k|\lambda, M)$. This is enough to hide our non-classical probability for pure states of single quantum systems, but not much more.

In our result, we saw that the use of standard measure theory when creating mixtures creates a problem because the convex space of classical distributions is significantly different from the convex space of quantum mixed states. For this case, ontological models are ``not quantum enough.''

In PBR, the issue is breaking a composite system into parts. They find  that if the probability measure on $\lambda$ is factorized, that is (\cite{PBR:2012}, eq. 5) 
\begin{equation}
	\int_{\Lambda} \dots \int_{\Lambda} p(k | \lambda_1, \cdots, \lambda_n) \mu_{x_1}(\lambda_1)\cdots\mu_{x_n}(\lambda_n) d\lambda_1 \cdots d\lambda_n ,
\end{equation}
there is a problem. The issue here is not the factorization. The issue here is breaking the composite system into parts, each with its own ontic state as in 
\begin{equation}
	\begin{aligned}
	\int_\Lambda &p(k | \lambda) d\lambda = \\
	&\int_{\Lambda_1} \dots \int_{\Lambda_n} p(k | \lambda_1, \cdots, \lambda_n) \mu(\lambda_1, \cdots, \lambda_n) d\lambda_1 \cdots d\lambda_n.
	\end{aligned}
\end{equation}
This assumes that a composite system is describable using a standard classical probability measure, which, in turn, allows us to express the joint probability distribution of observables of different subsystems. This, in general, will fail because it is ``not quantum enough.''

Therefore there are two different ways ontological models use classical measure theory: one in mixtures, highlighted by us; one in composite systems, highlighted by PBR. The use of classical measure theory is what drives the failure of the model.

The moral of the story is
\begin{quote}
	\textbf{In the context of quantum mechanics, standard probability measures are allowed \emph{only} over measurement outcomes/orthogonal states.}
\end{quote}
This, in fact, is exactly what we found in this paper using information theoretic considerations. In retrospect, as we said, it should not be at all surprising.

Note that HS do not claim that all interpretations must adhere to their framework, stating (\cite{Harrigan:2010}, p. 134), ``Therefore, to categorize any given interpretation of the quantum formalism as $\psi$-ontic or $\psi$-epistemic, it is first necessary to cast it into the mold of an ontological model. If an interpretation resists being so cast, then it cannot be fit into our categorization.'' The result is that a successful interpretation of quantum mechanics must resist being so cast.

@@

Our conclusion is that the PBR theorem finds that $\psi$-epistemic models are not sustainable not because there is a problem with $\psi$-epistemic models in particular, but rather there is an underlying problem in the HS definitions. Specifically, the use of standard probability theory in composite systems. If understood in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories, an argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.
